---
title: "Class 5: Data viz with ggplot"
author: "Junlin Ruan (PID: A17839687)"
format: gfm
---

Today we are exploring the **ggplot** package and how to make nice figures in R.

There are lots of ways to make figures and plot in R. These include:

- so called "base" R
- and add on packages like **ggplot2**

Here is a simple "base" R plot.

```{r}
head(cars)
```

We can simply pass to the 'plot()' function

```{r}
plot(cars)
```

> Key-point: Base R is quick but not so nice and simple looking in some folks' eyes.

Let's see how we can plot this with **ggplot2**...

1st I need to install this add-on package. For this we use the `install.packages()` function - **WE DO THIS IN THE CONSOLE, NOT our report**. This is a one time only deal.

2nd We need to load the package with `library()` function every time we want to use it.

```{r}
library(ggplot2)
ggplot(cars)
```

Every ggplot is composed of at least 3 layers:

- **data** (i.e. a data.frame with the things you want to plot),
- aesthetics **aes()** that map the columns of data to your plot features (i.e. aesthitics)
- geoms like **geom_point()** that srt how the plot appears

```{r}
ggplot(cars) +
  aes(x=speed, y=dist) +
  geom_point()
```
```{r}
hist(cars$speed)
```

> For simple "canned" graphs base R is quicker but as things get more custom and elaborate then ggplot wins out...

Let's add more layers to our ggplot

Add a line showing the relationship between x and y
Add a title
Add custom axis label "Speed (MPH)" and "Distance (ft)"
Change the theme...

```{r}
ggplot(cars) +
  aes(x=speed, y=dist) +
  geom_point() +
  geom_smooth(method="lm", se=FALSE)+
  labs(title="Silly plot of Speed vs Stoping distance", x="Speed (MPH)", y="Distance (ft)") +
  theme_bw()
```

## Going further

Read some gene expression data

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)

head(genes)
```

> Q1. How many genes are in this dataset?

```{r}
nrow(genes)
```
```{r}
ncol(genes)
```

> Q2. How many "up" regulated genes are there?

```{r}
sum(genes$State == "up")
```

A useful function for counting up occurrences of things in a vector is the `table()` function.


```{r}
table(genes$State)
```

Make a v1 figure

```{r}
p <- ggplot(genes) +
  aes(x=Condition1, 
      y=Condition2, 
      col=State) +
  geom_point()

p
```

```{r}
p + 
  scale_colour_manual(values=c("blue","gray","red"))+
  labs(title="Expression changes upon drug treatment", 
       x="Control (no drug)", 
       y="Treatment (with drug)") +
  theme_bw()
```

## More plotting

Read in the gapminder dataset

```{r}
# File location online
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"

gapminder <- read.delim(url)
```

Lets have a wee peak

```{r}
head(gapminder, 3)
```

> Q4. How many different country values are in this dataset?

```{r}
nrow(gapminder)
```

```{r}
length(table(gapminder$country))
```

> Q5. How many different continent values are in this dataset?

```{r}
unique(gapminder$continent)
```
```{r}
ggplot(gapminder) +
  aes(gdpPercap, lifeExp, col=continent) +
  geom_point()
```

```{r}
ggplot(gapminder) +
  aes(gdpPercap, lifeExp, col=continent, label=country) +
  geom_point() +
  geom_text()
```

I can use the **ggrepel** package to make more sensible labels here.

```{r}
library(ggrepel)

ggplot(gapminder) +
  aes(gdpPercap, lifeExp, col=continent, label=country) +
  geom_point() +
  geom_text_repel()
```

I want a separate pannel per continent

```{r}
ggplot(gapminder) +
  aes(gdpPercap, lifeExp, col=continent, label=country) +
  geom_point() +
  facet_wrap(~continent)
```

## Summary

The main advantages of ggplot over base R plotting are:

1. **Layered Grammar of Graphics**: ggplot uses a consistent, layered approach where you build plots by adding layers (data, aesthetics, geoms, themes, etc.), making complex plots easier to construct and modify [[1]](https://drive.google.com/file/d/1BYSWJLROqxA1YpuDhJkzUolhiZqiOOKg/view?usp=drivesdk), [[3]](https://drive.google.com/file/d/1tFqKg9_nhVMmKYfiM1CQKDS2PmPwLh8n/view?usp=drivesdk), [[5]](https://drive.google.com/file/d/15xXaaIcCWOc_x1gJLdySWOd_sfMXTiaw/view?usp=drivesdk).

2. **Sensible Defaults**: ggplot provides attractive default styles and layouts, so your plots look polished without extensive tweaking. Base R requires more manual adjustment for publication-quality visuals [[1]](https://drive.google.com/file/d/1BYSWJLROqxA1YpuDhJkzUolhiZqiOOKg/view?usp=drivesdk), [[3]](https://drive.google.com/file/d/1tFqKg9_nhVMmKYfiM1CQKDS2PmPwLh8n/view?usp=drivesdk), [[5]](https://drive.google.com/file/d/15xXaaIcCWOc_x1gJLdySWOd_sfMXTiaw/view?usp=drivesdk).

3. **Declarative Syntax**: You specify what you want to show (e.g., which variables map to which aesthetics), rather than how to draw each element. This makes code more readable and easier to maintain [[1]](https://drive.google.com/file/d/1BYSWJLROqxA1YpuDhJkzUolhiZqiOOKg/view?usp=drivesdk), [[3]](https://drive.google.com/file/d/1tFqKg9_nhVMmKYfiM1CQKDS2PmPwLh8n/view?usp=drivesdk), [[5]](https://drive.google.com/file/d/15xXaaIcCWOc_x1gJLdySWOd_sfMXTiaw/view?usp=drivesdk).

4. **Customization and Scalability**: ggplot makes it straightforward to add layers (like color, shape, smoothing lines, annotations) and to combine multiple plots. Base R can be fiddly and time-consuming for these tasks [[1]](https://drive.google.com/file/d/1BYSWJLROqxA1YpuDhJkzUolhiZqiOOKg/view?usp=drivesdk), [[3]](https://drive.google.com/file/d/1tFqKg9_nhVMmKYfiM1CQKDS2PmPwLh8n/view?usp=drivesdk), [[5]](https://drive.google.com/file/d/15xXaaIcCWOc_x1gJLdySWOd_sfMXTiaw/view?usp=drivesdk).

5. **Reproducibility**: ggplot code is modular and scriptable, making it easier to reproduce and share visualizations, especially in reports and collaborative projects [[1]](https://drive.google.com/file/d/1BYSWJLROqxA1YpuDhJkzUolhiZqiOOKg/view?usp=drivesdk), [[3]](https://drive.google.com/file/d/1tFqKg9_nhVMmKYfiM1CQKDS2PmPwLh8n/view?usp=drivesdk), [[5]](https://drive.google.com/file/d/15xXaaIcCWOc_x1gJLdySWOd_sfMXTiaw/view?usp=drivesdk).
