---
title: "Class10: Halloween mini project"
author: "Junlin Ruan (PID: A17839687)"
format: gfm
toc: TRUE
---

## Background

As it is nearly Halloween and the half way point in the quarter let's do a mini project that help us figure out the best candy!

Our data come from the 538 website and is available as a CSV file:

## Data Export

```{r}
candy <- read.csv("candy-data.txt", row.names = 1)
head(candy)
```

```{r}
library(flextable)
flextable::flextable(head(candy, 10))
```

> Q1. How many different candy types are in this dataset?

```{r}
nrow(candy)
```
```{r}
candy |>
  nrow()
```

```{r}
library(tidyverse)

candy %>%
  nrow()
```

> Q2. How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity)
```
My favorite winpercent
```{r}
library(dplyr)

candy |>
  filter(rownames(candy) == "Almond Joy") |>
  select(winpercent)
```

```{r}
candy["Twix", ]$winpercent
```

> Q3. What is your favorite candy in the dataset and what is it’s winpercent value?

```{r}
candy["Air Head", ]$winpercent
```

> Q4. What is the winpercent value for “Kit Kat”?

```{r}
candy["Kit Kat", ]$winpercent
```

> Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["Toostie Roll Snack Bars", ]$winpercent
```

## Quick overview of the dataset

```{r}
library("skimr")
skimr::skim(candy)
```

> Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

The winpercent is on a 0-100 scale the rest are 0-1 scale

> Q7. What do you think a zero and one represent for the candy$chocolate column?

That the candy does not contain chocolate

> Q8. Plot a histogram of winpercent values

```{r}
library(ggplot2)

ggplot(candy) +
  aes(winpercent) +
  geom_histogram(bins = 20)
```

> Q9. Is the distribution of winpercent values symmetrical?

```{r}
ggplot(candy) +
  aes(winpercent) +
  geom_density()
```

> Q10. Is the center of the distribution above or below 50%?

```{r}
mean(candy$winpercent)
```
```{r}
summary(candy$winpercent)
```

> Q11. On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
# 1. Find all chocolate candy in the data set
# 2. Find their winpercent values
# 3. Calculate the mean of these values

# 4-6. Do the same for fruity candy
# 7. Compare mean winpercents of chocolate vs fruity
# 8. Pick the highest as the winner

choc.inds <- candy$chocolate == 1
choc.win <- candy[choc.inds, ]$winpercent
choc.mean <- mean(choc.win)
choc.mean
```

```{r}
mean(candy[candy$chocolate==1,]$winpercent)
mean(candy[candy$fruity==1,]$winpercent)
```
```{r}
frut.ind <- candy$fruity==1
frut.win <- candy[frut.ind,]$winpercent
frut.mean <- mean(frut.win)
frut.mean
```

```{r}
candy |>
  filter(chocolate==1) |>
  select(winpercent) 
```

> Q12. Is this difference statistically significant?

```{r}
t.test(choc.win, frut.win)
```

## Overall Candy Ranking

> Q13. What are the five least liked candy types in this set?

```{r}
candy |>
  arrange(winpercent) |>
  head(5)
```

```{r}
x <- c(5,1,10,4)
#sort(x)
order(x)
#order(candy$winpercent)
```
```{r}
ord.ind <- order(candy$winpercent)
head(candy[ord.ind,], 5)
```


> Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
candy |>
  arrange(winpercent) |>
  tail(5)
```
```{r}
candy |>
  arrange(-winpercent) |>
  head(5)
```

> Q15. Make a first barplot of candy ranking based on winpercent values.

```{r}
ggplot(candy) +
  aes(winpercent, rownames(candy)) + 
  geom_col()
```

> Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
ggplot(candy) +
  aes(x = winpercent, 
      y = reorder(rownames(candy), winpercent)) + 
  geom_col()
```

Add some color based on the "type" of candy

```{r}
my_cols <- rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] <- "chocolate"
my_cols[as.logical(candy$fruity)] <- "pink"
my_cols[as.logical(candy$bar)] <- "brown"
my_cols
```

```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) 
```

> Q17. What is the worst ranked chocolate candy?

Sixlets
>Q18. What is the best ranked fruity candy?

Starburst

## Winpercent and Pricepercent

A plot with both variables/columns winpercent and pricepercent

```{r}
my_cols[as.logical(candy$fruity)] <- "red"
ggplot(candy) +
  aes(x = winpercent, 
      y = pricepercent, 
      label = rownames(candy)) + 
  geom_point(col=my_cols) +
  geom_text()
```

```{r}
library(ggrepel)
my_cols[as.logical(candy$fruity)] <- "red"
ggplot(candy) +
  aes(x = winpercent, 
      y = pricepercent, 
      label = rownames(candy)) + 
  geom_point(col=my_cols) +
  geom_text_repel(max.overlaps = 7)
```

> Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

Reese's Miniatures
> Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head(candy[ord, c(1, 11, 12)], n = 5)
```


## Exploring the correlation structure

Now that we’ve explored the dataset a little, we’ll see how the variables interact with one another. We’ll use correlation and view the results with the corrplot package to plot a correlation matrix.

```{r}
library(corrplot)
cij <- cor(candy)
corrplot(cij)
```

> Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

fruity and chocolate
> Q23. Similarly, what two variables are most positively correlated?

winpercent and chocolate

## Principal Component Analysis

The function to use is called `prcomp()` with an optional `scale=T/F` argument.

```{r}
pca <- prcomp(candy, scale = TRUE)
summary(pca)
```

Our main PCA result figure

```{r}
ggplot(pca$x) +
  aes(PC1, PC2, label = rownames(pca$x)) +
  geom_point(col = my_cols) +
  geom_text_repel(col = my_cols)
```

We should also examine the variable "loading" or contributions of the original variables to the new PCs

```{r}
ggplot(pca$rotation) +
  aes(PC1, rownames(pca$rotation)) +
  geom_col()
```

```{r}
p <- ggplot(pca$x) +
  aes(PC1, PC2, label = rownames(pca$x)) +
  geom_point(col = my_cols) +
  geom_text_repel(col = my_cols)
```

Interactive plots that can be zoomed on and "brushed" over can be made with the **plotly** package. Its output is interactive and will not render to PDF :-(

```{r}
library(plotly)

#ggplotly(p)
```

> Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

PC1 picked up "fruity", "hard", "pluribus" in the positive direction. It makes sense as it picks up candies that fall into these categories.